Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-27 curvature.tex
scatter svgs

\author{
	\IEEEauthorblockN{Christopher Whidden}
	\IEEEauthorblockA{Program in Computational Biology\\
	Fred Hutchinson Cancer Research Center\\
	Seattle, WA 98199\\
	Email: cwhidden@fredhutch.org}
	\and
	\IEEEauthorblockN{Frederick A Matsen IV}
	\IEEEauthorblockA{Program in Computational Biology\\
	Fred Hutchinson Cancer Research Center\\
	Seattle, WA 98199\\
	Email: matsen@fredhutch.org}
}

@@

\begin{abstract}
Statistical phylogenetics inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the SPR graph.
Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
This notion of curvature is especially relevant to stochastic search on the graph.
By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
This indicates significant structure of the graph; a greater understanding of this structure could ultimately lead to improved strategies for tree search.
\end{abstract}

>>>>

\author{
	\IEEEauthorblockN{Christopher Whidden}
	\IEEEauthorblockA{Program in Computational Biology\\
	Fred Hutchinson Cancer Research Center\\
	Seattle, WA 98109\\
	Email: cwhidden@fredhutch.org}
	\and
	\IEEEauthorblockN{Frederick A Matsen IV}
	\IEEEauthorblockA{Program in Computational Biology\\
	Fred Hutchinson Cancer Research Center\\
	Seattle, WA 98109\\
	Email: matsen@fredhutch.org}
}

@@

\begin{abstract}
Statistical phylogenetic inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the SPR graph.
Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
This notion of curvature is especially relevant to stochastic search on the graph.
By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
We confirm using simulation that degree and curvature have the expected impact on mean access time distributions.
This indicates significant structure of the graph beyond that which which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
\end{abstract}
